**Effective application of machine learning to a microbiome-based classification problem**

### 2020-01-21 submission/manuscript.Rmd
updates for rebuttal

ML models can be used to increase our understanding of the variation in the structure of existing data and in making predictions about new data. Researchers have used ML models to diagnose and understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases, obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018; @thaiss_persistent_2016; @dadkhah_gut_2019; @flemer_oral_2018; @montassier_pretreatment_2016; @ai_systematic_2017; @dai_multi-cohort_2018; @mossotto_classification_2017]. The task of diagnosing an individual relies on a rigorously validated model. However, there are common methodological and reporting problems that arise when applying ML to such data, that need to be addressed for the field to progress. These problems include a lack of transparency in which methods are used and how these methods are implemented; evaluating models without separate held-out test data; unreported variation between the predictive performance on different folds of cross-validation; and unreported variation between cross-validation and testing performances. Though the microbiome field is making progress to avoid some of these pitfalls including validating their models on independent datasets [@wong_quantitation_2017; @mossotto_classification_2017; @sze_looking_2016] and introducing ways to better use ML tools  [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019], more work is needed to improve reproducibility further and minimize overestimating for model performance.

@@

In this study, we provided steps toward standardization of machine learning methods for microbiome studies which are often poorly documented and executed. To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing seven modeling approaches with the same dataset and pipeline. We built three linear models with different forms of regularization: L2-regularized logistic regression and L1 and L2-regularized support vector machines (SVM) with a linear kernel. We also trained four non-linear models: SVM with radial basis function kernel, a decision tree, random forest, and gradient boosted trees. We compared their predictive performance, interpretability, and training time. To demonstrate the performance of these modeling approaches and our pipeline on a case study, we used a data from a previously published study that sought to classify individuals as having healthy colons or colonic lesions based on the 16S rRNA gene sequences collected from fecal samples [@baxter_dna_2016]. This dataset was selected because it is a relatively large collection of individuals (N=490) connected to a clinically significant disease where there is ample evidence that the disease is driven by variation in the microbiome [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011]. With this dataset, we developed an ML pipeline that can be used in many different scenarios for training and evaluating models. This framework can be easily applied to other host-associated and environmental microbiome datasets. We hope that this study ...

@@

First, we randomly split the data into training and test sets so that the training set consisted of 80% of the full dataset, while the test set was composed of the remaining 20% [Figure 1]. To maintain the distribution of controls and cases found in the full dataset, we performed stratified splits. For example, our full dataset included 490 individuals. Of these, 261 had healthy colons (53%) and 229 had a screen relevant neoplasia (SRN; 46.7%). A training set included 393 individuals, of which 209 had an SRN (53%), while the test set was composed of 97 individuals, of which 52 had an SRN (54%). The training data were used to build and select the models, and the test set was used for evaluating the model.

We trained seven different models using the training data [Table 1]. We focused on different classification algorithms and regularization methods. Regularization helps to prevent overfitting by penalizing a model that fits the training data too well [@Ng:2004:FSL:1015330.1015435]. For regularized logistic regression and SVM with a linear kernel, we used L2-regularization to keep all potentially important features. For comparison, we also trained an L1-regularized SVM with a linear kernel. L1-regularization on microbiome data led to a sparser solution (i.e., forced many coefficients to zero). To explore the potential for non-linear relationships among features to improve classification, we trained tree-based models including, a decision tree, a random forest, and gradient boosted trees (XGBoost) and an SVM with a non-linear kernel.

@@

Second, to analyze non-linear models, we interpreted the feature importance using permutation importance [@10.1093/bioinformatics/btq134]. Whereas the absolute feature weights were determined from the trained models, here we measured importance using the held-out test data. Permutation importance analysis is a post hoc explanation of the model, in which we randomly permuted groups of perfectly correlated features together and other features individually across the two groups in the held-out test data. We then calculated how much the predictive performance of the model (i.e., testing AUROC values) decreased when each OTU or group of OTUs was randomly permuted. We ranked the OTUs based on how much the median testing AUROC decreased when it was permuted; the OTU with the largest decrease ranked highest [Figure 4]. Among the twenty OTUs with the largest impact, there was only one OTU (OTU 822) that was shared among all of the models; however, we found three OTUs (OTUs 58, 110, 367) that were important in each of the tree-based models. Similarly, the random forest and XGBoost models shared four of the most important OTUs (OTUs 2, 12, 361, 477). Permutation analysis results also revealed that with the exception of the decision tree model, removal of any individual OTU had minimal impact on model performance. For example, if OTU 367 was permuted across the samples in the decision tree model, the median AUROC dropped from `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` to `r paste(round(DT$imp, 3))`. In contrast, if the same OTU was permuted in the random forest model, the AUROC only dropped from `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` to `r paste(format(round(RF$imp, 3), nsmall=3))`, which indicated high degree of collinearity in the dataset. Permutation analysis allowed us to gauge the importance of each OTU in non-linear models and partially account for collinearity by grouping correlated OTUs to determine their impact as a group.

To further highlight the differences between the two interpretation methods, we used permutation importance to interpret the linear models [Figure S6]. When we analyzed the L1-regularized SVM with linear kernel model using feature rankings based on weights [Figure 3] and permutation importance [Figure S6], 17 of the 20 top OTUs (e.g., OTU 609, 822, 1239) were deemed important by both interpretation methods. Similarly, for the L2-regularized SVM and L2-regularized logistic regression, 9 and 12 OTUs, respectively, were shared among the two interpretation methods. These results indicate that both methods are consistent in selecting the most important OTUs.

@@

We highlighted the importance of model interpretation to gain greater biological insights into microbiota-associated diseases. In this study, we showcased two different interpretation methods: ranking each OTU by (i) their absolute weights in the trained models and (ii) their impact on the predictive performance based on permutation importance. Human-associated microbial communities have complex correlation structures that create collinearity in the datasets. This can hinder our ability to reliably interpret models because the feature weights of correlated OTUs are influenced by one another [@dormann_collinearity:_2013]. To capture all important features, once we identify highly ranked OTUs, we should review their relationships with other OTUs. These relationships will help us generate new hypotheses about the ecology of the disease and test them with follow-up experiments. When we used permutation importance, we partially accounted for collinearity by grouping correlated OTUs to determine their impact as a group. We grouped OTUs that had a perfect correlation with each other; however, we could reduce the correlation threshold to further investigate the relationships among correlated features. We are choosing to leave the decision to define correlation and setting up correlations thresholds to the researchers for their own microbiome questions. However, we would like to urge them to investigate the correlation structures of the data to avoid misinterpreting the models. This is likely to be a particular problem with shotgun metagenomic datasets where collinearity will be more pronounced due to many genes being correlated with one another because they come from the same chromosome. True causal mechanisms (e.g. role of microbiome in colorectal cancer) cannot be explained solely by the highest performing machine learning model [@li_accurate_2020]. To identify the true underlying microbial factors of a disease, it is crucial to follow up on any correlation analyses with further hypothesis testing and experimentation for biological validation.

@@

Random forest & \makecell[l]{A classifier that is an ensembe of decision trees \\ that grows randomly with subsampled data.} & Non-linear \\ \hline

@@

**Figure 4. Interpretation of the non-linear ML models.** (A) SVM with radial basis kernel, (B) decision tree, (C) random forest, and (D) XGBoost feature importances were explained using permutation importance on the held-out test data set. The gray rectangle and the dashed line show the IQR range and median of the base testing AUROC without any permutation. The 20 OTUs that caused the largest decrease in the AUROC when permuted are reported here. The colors of the box plots represent the OTUs that were shared among the different models; yellow were OTUs that were shared among all the non-linear models, salmon were OTUs that were shared among the tree-based models, green were the OTUs shared among SVM with radial basis kernel, decision tree and XGBoost, pink were the OTUs shared among SVM with radial basis kernel and XGBoost only, red were the OTUs shared among random forest and XGBoost only and blue were the OTUs shared among decision tree and random forest only. For all of the tree-based models, a *Peptostreptococcus* species (OTU00367) had the largest impact on predictive performance.
\newpage
\includegraphics{Figure_5.png}

@@

\makecell[l]{Study \\ cohort} & \makecell[l]{Test data resampled to remove \\ class imbalance (e.g., test data \\ resampled to have an equal \\ number of patients with carcinomas \\ as patients with healthy colons, \\ which does not reflect reality.)}  & \makecell[l]{Test data are reflective \\ of the population which \\ the model will be applied.} & \makecell[l]{Model tested on multiple \\ cohorts with potentially \\ different class balances.} \\ \hline

@@

\makecell[l]{Evaluation \\ metrics} & \makecell[l]{Reported performance according to \\ a metric that is not appropriate \\ for intended application  (e.g. when \\  predicting rare outcome, accuracy \\ metric is not reliable).} & \makecell[l]{Reported performance in \\ terms of a metric that \\ is appropriate for intended \\ application and includes \\ confidence intervals.} & \makecell[l]{Reported multiple metrics \\ with confidence intervals.} \\ \hline

@@

**Figure S3. Histogram of AUROC differences between L2-regularized logistic regression and random forest for each of the hundred data-splits.** This histogram shows the number of data-splits in each bin. The percentage of dataplits where the difference between random forest and L2-regularized logistic regression AUROC values was higher than or equal to 0 were 0.75, lower than or equal to 0 were 0.25. The vertical red line highlights the bins where there AUROC difference between the two model is 0. The p-value was manually calculated using the sampling distribution of the test statistic under the null hypothesis. We are testing how often random forest performs more accurately than L2-regularized logistic regression. We are assuming that the distribution of the differences between the AUROC values of random forest and L2 logistic regression is symmetric about 0, therefore we are calculating the p-value for a double-tail event.
\newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S4.png}

@@

**Figure S5. NMDS ordination with Bray-Curtis distances.**

@@

**Figure S6. Interpretation of the linear ML models with permutation importance.** (A) L1-regularized SVM with linear kernel, (B) L2-regularized SVM with linear kernel, and (C) L2-regularized logistic regression were interpreted using permutation importance using held-out test set.

>>>>

ML models can be used to increase our understanding of the variation in the structure of existing data and in making predictions about new data. Researchers have used ML models to diagnose and understand the ecological basis of diseases such as liver cirrhosis, colorectal cancer, inflammatory bowel diseases, obesity, and type 2 diabetes [@zeller_potential_2014; @zackular_human_2014; @baxter_dna_2016; @baxter_microbiota-based_2016; @hale_shifts_2017; @pasolli_machine_2016; @sze_looking_2016; @walters_meta-analyses_2014; @vazquez-baeza_guiding_2018; @qin_alterations_2014; @geman_deep_2018; @thaiss_persistent_2016; @dadkhah_gut_2019; @flemer_oral_2018; @montassier_pretreatment_2016; @ai_systematic_2017; @dai_multi-cohort_2018; @mossotto_classification_2017]. The task of diagnosing an individual relies on a rigorously validated model. However, there are common methodological and reporting problems that arise when applying ML to such data, that need to be addressed for the field to progress. These problems include a lack of transparency in which methods are used and how these methods are implemented; evaluating models without separate held-out test data; unreported variation between the predictive performance on different folds of cross-validation; and unreported variation between cross-validation and testing performances. Though the microbiome field is making progress to avoid some of these pitfalls including validating their models on independent datasets [@wong_quantitation_2017; @mossotto_classification_2017; @sze_looking_2016] and introducing accessible and open-source ML tools  [@statnikov_comprehensive_2013; @knights_supervised_2011; @wirbel_meta-analysis_2019; @vangay_microbiome_2019], more work is needed to improve reproducibility further and minimize overestimating for model performance.

@@

In this study, we provided steps toward standardization of machine learning methods for microbiome studies which are often poorly documented and executed. To showcase a rigorous ML pipeline and to shed light on how ML model selection can affect modeling results, we performed an empirical analysis comparing the predictive performance, interpretability and training times of seven modeling approaches with the same dataset and pipeline. We built three linear models with different forms of regularization: L2-regularized logistic regression and L1 and L2-regularized support vector machines (SVM) with a linear kernel. We also trained four non-linear models: SVM with radial basis function kernel, a decision tree, random forest, and gradient boosted trees. We compared their predictive performance, interpretability, and training time. To demonstrate the performance of these modeling approaches and our pipeline on a case study, we used a data from a previously published study that sought to classify individuals as having healthy colons or colonic lesions based on the 16S rRNA gene sequences collected from fecal samples [@baxter_dna_2016]. This dataset was selected because it is a relatively large collection of individuals (N=490) connected to a clinically significant disease where there is ample evidence that the disease is driven by variation in the microbiome [@baxter_microbiota-based_2016; @baxter_dna_2016; @zeller_potential_2014; @knights_human-associated_2011]. With this dataset, we developed an ML pipeline that can be used in many different scenarios for training and evaluating models. This framework can be easily applied to other host-associated and environmental microbiome datasets. We also provided an aspirational rubric for evaluating the rigor of ML practices applied to microbiome data [Figure S1] to urge the audience to be diligent with with their study design, model selection, development, evaluation and interpretation.

@@

First, we randomly split the data into training and test sets so that the training set consisted of 80% of the full dataset, while the test set was composed of the remaining 20% [Figure 1]. To maintain the distribution of controls and cases found in the full dataset, we performed stratified splits. For example, our full dataset included 490 individuals. Of these, 261 had healthy colons (53%) and 229 had a screen relevant neoplasia (SRN; 46.7%). A training set included 393 individuals, of which 209 had an SRN (53%), while the test set was composed of 97 individuals, of which 52 had an SRN (54%). The training data were used to build and select the models, and the test set was used for evaluating the model. We trained seven different models using the training data [Table 1].

@@

Second, to analyze non-linear models, we interpreted the feature importance using permutation importance [@10.1093/bioinformatics/btq134]. Whereas the absolute feature weights were determined from the trained models, here we measured importance using the held-out test data. Permutation importance analysis is a post hoc explanation of the model, in which we randomly permuted groups of perfectly correlated features together and other features individually across the two groups in the held-out test data [Figure S6]. We then calculated how much the predictive performance of the model (i.e., testing AUROC values) decreased when each OTU or group of OTUs was randomly permuted. We ranked the OTUs based on how much the median testing AUROC decreased when it was permuted; the OTU with the largest decrease ranked highest [Figure 4]. Among the twenty OTUs with the largest impact, there was only one OTU (OTU 822) that was shared among all of the models; however, we found three OTUs (OTUs 58, 110, 367) that were important in each of the tree-based models. Similarly, the random forest and XGBoost models shared four of the most important OTUs (OTUs 2, 12, 361, 477). Permutation analysis results also revealed that with the exception of the decision tree model, removal of any individual OTU had minimal impact on model performance. For example, if OTU 367 was permuted across the samples in the decision tree model, the median AUROC dropped from `r paste(round(median((all %>% filter(model=="Decision_Tree"))$test_aucs), 3))` to `r paste(round(DT$imp, 3))`. In contrast, if the same OTU was permuted in the random forest model, the AUROC only dropped from `r paste(round(median((all %>% filter(model=="Random_Forest"))$test_aucs), 3))` to `r paste(format(round(RF$imp, 3), nsmall=3))`, which indicated high degree of collinearity in the dataset. Permutation analysis allowed us to gauge the importance of each OTU in non-linear models and partially account for collinearity by grouping correlated OTUs to determine their impact as a group.

To further highlight the differences between the two interpretation methods, we used permutation importance to interpret the linear models [Figure S7]. When we analyzed the L1-regularized SVM with linear kernel model using feature rankings based on weights [Figure 3] and permutation importance [Figure S6], 17 of the 20 top OTUs (e.g., OTU 609, 822, 1239) were deemed important by both interpretation methods. Similarly, for the L2-regularized SVM and L2-regularized logistic regression, 9 and 12 OTUs, respectively, were shared among the two interpretation methods. These results indicate that both methods are consistent in selecting the most important OTUs.

@@

We highlighted the importance of model interpretation to gain greater biological insights into microbiota-associated diseases. In this study, we showcased two different interpretation methods: ranking each OTU by (i) their absolute weights in the trained models and (ii) their impact on the predictive performance based on permutation importance. Human-associated microbial communities have complex correlation structures that create collinearity in the datasets. This can hinder our ability to reliably interpret models because the feature weights of correlated OTUs are influenced by one another [@dormann_collinearity:_2013]. To capture all important features, once we identify highly ranked OTUs, we should review their relationships with other OTUs. These relationships will help us generate new hypotheses about the ecology of the disease and test them with follow-up experiments. When we used permutation importance, we partially accounted for collinearity by grouping correlated OTUs to determine their impact as a group. We grouped OTUs that had a perfect correlation with each other; however, we could reduce the correlation threshold to further investigate the relationships among correlated features. We are choosing to leave the decision to define correlation and setting up correlations thresholds to the researchers for their own microbiome questions. However, we would like to urge them to investigate the correlation structures of the data to avoid misinterpreting the models. This is likely to be a particular problem with shotgun metagenomic datasets where collinearity will be more pronounced due to many genes being correlated with one another because they come from the same chromosome. True causal mechanisms (e.g., role of microbiome in colorectal cancer) cannot be explained solely by the highest performing machine learning model [@li_accurate_2020]. To identify the true underlying microbial factors of a disease, it is crucial to follow up on any correlation analyses with further hypothesis testing and experimentation for biological validation.

@@

Random forest & \makecell[l]{A classifier that is an ensemble of decision trees \\ that grows randomly with subsampled data.} & Non-linear \\ \hline

@@

**Figure 4. Interpretation of the non-linear ML models.** (A) SVM with radial basis kernel, (B) decision tree, (C) random forest, and (D) XGBoost feature importances were explained using permutation importance on the held-out test data set. The gray rectangle and the dashed line show the IQR range and median of the base testing AUROC without any permutation. The 20 OTUs that caused the largest decrease in the AUROC when permuted are reported here. The colors of the box plots represent the OTUs that were shared among the different models; yellow were OTUs that were shared among all the non-linear models, green were OTUs that were shared among the tree-based models, green were the OTUs shared among SVM with radial basis kernel, decision tree and XGBoost, pink were the OTUs shared among SVM with radial basis kernel and XGBoost only, red were the OTUs shared among random forest and XGBoost only and blue were the OTUs shared among decision tree and random forest only. For all of the tree-based models, a *Peptostreptococcus* species (OTU00367) had the largest impact on predictive performance.
\newpage
\includegraphics{Figure_5.png}

@@

\makecell[l]{Study \\ cohort} & \makecell[l]{Test data resampled to remove \\ class imbalance (e.g., test data \\ resampled to have an equal \\ number of patients with carcinomas \\ as patients with healthy colons, \\ which does not reflect reality.)}  & \makecell[l]{Test data are reflective \\ of the population to which \\ the model will be applied.} & \makecell[l]{Model tested on multiple \\ cohorts with potentially \\ different class balances.} \\ \hline

@@

\makecell[l]{Evaluation \\ metrics} & \makecell[l]{Reported performance according to \\ a metric that is not appropriate \\ for intended application  (e.g., when \\  predicting rare outcome, accuracy \\ metric is not reliable).} & \makecell[l]{Reported performance in \\ terms of a metric that \\ is appropriate for intended \\ application and includes \\ confidence intervals.} & \makecell[l]{Reported multiple metrics \\ with confidence intervals.} \\ \hline

@@

**Figure S3. Histogram of AUROC differences between L2-regularized logistic regression and random forest for each of the hundred data-splits.**  In 75% of data-splits, the AUROC of random forest was greater than that of L2-regularized logistic regression. The p-value was manually calculated using the sampling distribution of the test statistic under the null hypothesis. We tested how often random forest performed more accurately than L2-regularized logistic regression. The null hypothesis is that the distribution of the difference between the AUROC values of random forest and L2 logistic regression is symmetric about 0, therefore the p-value was calculated for a double-tail event.
\newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S4.png}

@@

**Figure S5. NMDS ordination of Bray-Curtis distances.** NMDS ordination relating the community structures of the fecal microbiota from 490 patients (261 patients with normal colonoscopies and 229 patients who have screen relevant neoplasias; SRNs).

@@

**Figure S6. Permutation importance analysis.** Permutation importance analysis measures the decrease in the predictive performance of the model after we permute (A) a feature’s or (B) a group of correlated features' values, which breaks the relationship between the feature and the diagnosis.

\newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S7.png}

**Figure S7. Interpretation of the linear ML models with permutation importance.** (A) L1-regularized SVM with linear kernel, (B) L2-regularized SVM with linear kernel, and (C) L2-regularized logistic regression were interpreted using permutation importance using held-out test set.

\newpage
\includegraphics[height=17.5cm, width=13cm]{Figure_S7.png}

**Figure S8. Interpretation of the linear ML models with permutation importance.** (A) L1-regularized SVM with linear kernel, (B) L2-regularized SVM with linear kernel, and (C) L2-regularized logistic regression were interpreted using permutation importance using held-out test set.
